Archaic introgression and the distribution of shared variation under stabilizing selection

### 2024-08-20 doc/manuscript.tex
careful reading

    Many phenotypic traits are under stabilizing selection, which maintains a
    population's mean phenotypic value near some optimum. The dynamics of
    traits and trait architectures under stabilizing selection have bene
    extensively studied for single populations at steady state. However,
    natural populations are seldom at steady state. Instead, they are often
    structured in some way, and admixture and introgression events may be
    common, including over human evolutionary history. Because stabilizing
    selection results in selection against the minor allele at a
    trait-affecting locus, alleles from the minor parental ancestry will be
    selected against after admixture. Here, we show that the site-frequency
    spectrum can be used to model the genetic architecture of such traits,
    allowing for the study of trait architecture dynamics in complex
    multi-population settings. We use a simple deterministic two-locus model to
    predict the reduction of introgressed ancestry around trait-contributing
    loci. From this and individual-based simulations, we show that
    introgressed-ancestry deserts are enriched around such loci. When
    introgression between two diverged populations occurs in both directions,
    as has been inferred between humans and Neanderthals, the locations of
    introgressed-ancestry deserts will tend to be shared across populations. We
    argue that stabilizing selection for shared phenotypic optima may explain
    recent observations in which regions of depleted human-introgressed
    ancestry in the Neanderthal genome overlap with Neanderthal-ancestry
    deserts in humans.

@@

Once introduced through admixture, introgressed alleles may be selected for or
against. Some introgressed haplotypes were likely positively selected in modern
\emph{Homo sapiens} (here, ``humans'') \citep{huerta2014altitude,
racimo2017signatures, enard2018evidence, gower2021detecting}, possibly due to
locally adaptive variation that provided fitness advantages as humans
encountered novel environments. Despite some cases of adaptive introgression,
most introduced alleles were likely to have been selected against in humans
\citep{harris2016genetic, juric2016strength, veller2023recombination}. Since
Neanderthal and Denisovan population sizes were relatively small for hundreds
of thousands of years, theory predicts they would have accumulated deleterious
variation at an increased rate. Introgressed haplotypes loaded with more
deleterious mutations would have been rapidly removed by selection after
admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
humans shows a reduction of Neanderthal-related ancestry in coding and
regulatory regions \citep{petr2019limits, telis2020selection,
yermakovich2023long}. These ``deserts'' of Neanderthal ancestry support the
hypothesis that introgressed functional alleles were selected against
\citep{sankararaman2014genomic, sankararaman2016combined}.

There is growing genetic evidence that humans reciprocally contributed to
Neanderthal genomes \citep{kuhlwilm2016ancient, hubisz2020mapping,
harris2023diverse, li2024recurrent}. This gene flow occurred tens to hundreds
of thousands of years prior to Neanderthal introgression in humans during the
global dispersal of modern humans around 60 ka. This is supported by early
\emph{H. sapiens} outside of African around 200--100 ka \citep{schwarcz1988esr,
grun2005u, harvati2019apidima, beyer2021climatic}, potentially overlapping with
Neanderthals and providing opportunities for early contacts. While estimates of
the genomic contribution of early humans to Neanderthals vary, around $6\%$ of
later Neanderthal genomes may trace through this admixture event
\citep{harris2023diverse}. Under a ``load'' model, if human-related haplotypes
carried fewer deleterious alleles due to their larger long-term effective
population size, human-introgressed DNA would have been favored in Neanderthal
genomes. The replacement of Neanderthal mitochondrial and Y chromosomes by
early human haplotypes appears to support this model of post-admixture positive
selection in the Neanderthal lineage \citep{posth2017deeply,
petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (that is, due to differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. These models, founded in
population genetics theory, rarely take into account selection operating on
phenotypic traits or the relationship between genetic and phenotypic variation.
Many phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence, sella2019thinking}, including gene regulation
\citep{gilad2006natural, hodgins2015gene, price2022detecting}. Because some of
the strongest signals of selection against Neanderthal-introduced alleles are
in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
quantitative traits may be particularly relevant to the dynamics of functional
genetic variation after introgression among hominins.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/stab_sel.pdf}
    \caption{
        \textbf{Additive genetic variance under stabilizing selection.}
        (A) Stabilizing selection acts to maintain phenotypic values of
        individuals in the population near some optimum. Throughout, we assume
        a Gaussian fitness function.
        (B) With low mutational variance ($V_M$), the expected additive
        genetic variance ($V_G$) is proportional to the
        population-scaled mutation rate. When $V_M$ is large so that
        mutational effects can be strong, $V_G$ is independent of the mutation
        rate. The stochastic house-of-cards model (Eq.~\ref{eq:SHC})
        interpolates these regimes, assuming steady-state dynamics
        \citep{burger1989much}. Expected $V_G$ computed using an SFS approach
        \citep[developed here using \moments,][]{jouganous2017inferring}
        matches simulations assuming no linkage between loci affecting the
        trait, which align closely with the stochastic house-of-cards
        approximation.
        Here $V_S=1$, $N_e=10{,}000$, the mutation rate
        $\mu=0.01$ (per haploid), and mutation effects are drawn from a normal
        distribution with mean zero and variance $V_M$.
    }
    \label{fig:stab-sel}
\end{figure}

In diverged populations, the genetic variation contributing to a trait under
stabilizing selection has a higher rate of turnover compared to neutral
evolution \citep{yair2022population}. We therefore expect a rapid divergence of
trait architectures and an accumulation of fixed differences in the human and
Neanderthal lineages at trait-contributing loci, even when the mean phenotype
in each population remains close to the same trait optimum. When a derived
allele at high frequency in one population is introduced to another population
in which it was previously absent, it will be at low frequency (if the
admixture proportion is low) and subsequently selected against. Likewise, if
the ancestral allele is reintroduced to a population fixed for a derived
allele, the ancestral allele will be at low frequency and will be selected
against. Selection acts against the introgressed allele in either case, whether
it is ancestral or derived and regardless of the historical relative sizes of
the populations involved. In concurrent work, \citet{veller2024stabilizing}
demonstrate this effect by deriving the expected decay of minor parental
ancestry under stabilizing selection after admixture. This prediction contrasts
with the population-genetics load model, in which haplotypes with fewer
deleterious variants (such as those from the population with larger historical
size) are favored after introgression in either direction.

In this article, we show that admixture between diverged populations results in
selection against the minor parental ancestry at loci contributing to a trait
under stabilizing selection. We develop a numerical approach based on the
site-frequency spectrum to predict genetic variation under complex demographic
scenarios, and which we use to partition predicted trait heritability by
introgressed and non-introgressed variation. Using simulations with linkage, we
demonstrate that deserts of introgressed ancestry form around
trait-contributing loci. When gene flow occurs bidirectionally, such deserts
will tend to overlap in location across populations. We argue that stabilizing
selection on shared trait optima may explain the overlap of introgressed
ancestry deserts in human and Neanderthal genomes after reciprocal
introgression, as has been recently observed \citep{harris2023diverse}.

@@

We consider a polygenic trait for which an individual's additive genetic value
is the sum over all effects of alleles in their genome. For individual $i$,
\(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size of the derived
allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
locus (i.e., the number of derived alleles they carry at that locus). With
linkage equilibrium between trait-affecting loci, the expected additive genetic
variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
frequency at locus $l$. We ignore dominance and epistatic effects (often argued
to be a reasonable modeling choice, e.g., \citet{barton2002understanding,
zhang2002pleiotropic}), so that \(V_G = V_A\). We further ignore environmental
effects and noise \citep[under the assumption of additive environmental
effects, as in][]{lande1975maintenance, simons2018population}, so that only
genetic effects are considered.

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$, typically set to zero, and we assume a Gaussian fitness function
(Figure~\ref{fig:stab-sel}A) so that relative fitness (relative to an
individual with optimal phenotype) is given by \(f(G_i | O,
V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $\frac{1}{V_S}$ is interpreted as the
strength of selection on the trait, so larger $V_S$ implies weaker selection.
For a population with mean phenotype at or very close to the optimum, the
mean fitness of the population [assuming a normal distribution of phenotypic
values in the population, \citep{turelli1994genetic, urban2013asymmetric}] is
\[\bar{w} \approx \int_{-\infty}^\infty
f(G | O, V_S) \mathcal{N}(O, V_G) dG =
\left(\frac{V_S}{V_S+V_G}\right)^{1/2},\]
so that as the genetic variance increases, population mean fitness decreases.

@@

In our case, the selection coefficient $s$ depends on the strength of selection
on the trait $V_S$ and the effect size $a$, as well as the frequency of the
allele, so that
\begin{equation}
    s\approx -\frac{a^2(1-2p)}{2(V_S+V_G)} 
    = \frac{\left(p-\frac{1}{2}\right)a^2}{V_S + V_G}
    \approx \frac{\left(p-\frac{1}{2}\right)a^2}{V_S},
\end{equation}
when $V_G \ll V_S$. This model assumes linkage equilibrium between
trait-contributing loci. We note that LD is expected to be generated between
loci even in the absence of linkage \citep{bulmer1971effect}, and
\citet{negm2024effect} recently showed how to correct for its effect on the
selection dynamics of a focal allele.

@@

We extended the \moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017inferring} to include underdominance with given
selection coefficient $s$ (Appendix~\ref{sec:moments-underdominance}). The
contribution of alleles with effect size $a$ to the total genetic variance
$V_G$ is found by computing the expected pairwise diversity from the SFS (with
sample size $n$, denoted $\Phi_n$), as
\[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1} \frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\]
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
assuming a normal distribution of effect sizes for new mutations, the total
genetic variance is
\[V_G \approx \int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
$V_G$ can be computed periodically in this way to predict trajectories in
non-equilibrium settings (Figure~\ref{fig:one-pop}).

If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
$s=(p-1/2)a^2/V_S$ leads to deflated estimates of additive genetic variance
(Figure~\ref{fig:supp-one-pop}). When mutation rates are large so that $V_G$ is
not small compared to $V_S$, using $s=(p-1/2)a^2/(V_S+V_G)$ provides estimates
of $V_G$ that closely match simulations assuming free recombination between
loci (Figures~\ref{fig:toy-admixture} and
\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}). Because $V_G$ can vary over
time, this means that $s$ is no longer constant and can change due to factors
such as non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Population size changes and additive genetic variance.}
        We used \moments to track expected genetic variance of a trait
        under stabilizing selection as the population changes size. (A) The
        population goes through a 10-fold size reduction, followed by a recovery.
        (B) All mutations have effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
        probability of being trait increasing or decreasing. Depending on the
        selection coefficient ($s\approx (p-1/2)a^2/V_S$) compared to $N_e$, $V_G$
        could increase or decrease after a sudden population size change.
        Predictions using \moments match simulations with free recombination
        between trait-affecting loci
        (Figures~\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}).
    }
    \label{fig:one-pop}
\end{figure}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model (which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence) a population of size $N_e=10{,}000$ splits into two, one remaining
size $10{,}000$ and the other shrinking to size $1{,}000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{fig:toy-admixture}A,B).

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/reciprocal_admixture.pdf}
    \caption{
        \textbf{Genetic variance of a trait under stabilizing selection with
        multi-population demography.}
        (A,B) Two simple demographic models, in which Deme 1 has size 10,000
        and Deme 2 has size 1,000. In each scenario, we allow $5\%$ admixture
        from one deme to the other after begin isolated for 20,000 ($2N_e$)
        generations.
        (C,D) We compared predicted (additive) $V_G$ from the site-frequency
        spectra (using \moments) to simulations
        without linkage between trait-affecting loci. In Deme 2, $V_G$ decreases
        after divergence and population size reduction. After admixture in both
        directions, $V_G$ increases and then decays again toward levels at steady
        state. Predictions from \moments closely match observed $V_G$ in
        simulations.
        Here, mutations were drawn from a normal distribution of fitnesses
        with $V_M=0.0025$.
        Other parameters: $\mu=0.025$, $V_S=1$, and the optimal phenotype remained
        the same in each population.
        (E,F) Using \moments, we partitioned additive contributions to $V_G$ by
        mutations that were previously segregating in the focal population,
        introgressed variants (either derived or reintroduced ancestral alleles),
        and by new mutations since the time of admixture. In both demographic
        scenarios, introgressed variants contribute a substantial proportion to
        $V_G$, though it is primarily composed of introduced derived alleles when
        admixture is from the small to the large population, while primarily
        reintroduced ancestral alleles under the reverse direction of gene flow.
        In both cases, $V_G$ is increasingly due to new mutations and the genetic
        architecture of the trait turns over with time since admixture.
    }
    \label{fig:toy-admixture}
\end{figure}

@@

We compared our \moments-based predictions for $V_G$ in non-equilibrium
settings to simulations assuming linkage equilibrium as well as
individual-based simulations with recombination (next section). For both,
mutations occur at rate $\mu$ per haploid genome copy per generation, with
effects drawn from some distribution of mutation effects.

@@

We typically expect genetic variance to increase after introgression. The
amount that genetic variance increases depends on the allelic differences
accumulated between populations and the effects of those alleles. Assuming
linkage equilibrium, and ignoring dominance and epistasis, \(V_G=\sum_l
2p_l(1-p_l)a_l^2\). After admixture, with proportion $f$ contributed by the
population labeled 0 and $1-f$ by population 1, \(p_l = fp_{l,0} +
(1-f)p_{l,1}\). Plugging into the expression for $V_G$ and after a bit of
algebra (Appendix~\ref{sec:VG-admixture}), we can write the expected genetic
variance directly after admixture as
\[V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\]
where $F_2 = (p_0 - p_1)^2$ is the squared difference in allele frequencies at
a locus \citep{peter2016admixture}. This result is known
\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
to that in the source populations weighted by their contributions, plus a term
that depends on the divergence at trait-affecting loci between the populations
weighted by the quadratic factor $2f(1-f)$. Notably, $V_G$ is expected to
increase only in the second generation after admixture, so that selection
against introgressed alleles is not immediate \citep{veller2024stabilizing}.
This effect is not captured by this expression.

$F_2$ at a given locus depends on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on demographic history. In this case,
\begin{align} \label{eq:VG-admixture}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    \nonumber
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
\end{align}
where $L$ is the number of trait affecting loci and assuming $\E[a]=0$.

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/h2-per-SNP.pdf}
    \caption{
        \textbf{Allele contributions to heritability under human-Neanderthal
        reciprocal introgression.}
        (A) A simple model of divergence and admixture between humans and
        Neanderthals. Using \moments, we computed predicted $V_G$ over time,
        partitioned by variation that was introgressed vs. non-introgressed
        (Figure~\ref{fig:human-neand-VG}).
        (B-E) Predicted per-SNP contributions to genetic variance ($h^2$ per
        SNP) is plotted over the 50 thousand years following introgression.
        For non-introgressed variants, we also plot $h^2$ per SNP weighted by
        allele frequencies matching those of introgressed variants. These
        are shown for (B,C) Neanderthal-to-human introgression 50 kya,
        (D,E) human-to-Neanderthal introgression 250 kya, (B,D) $V_M=0.0025$,
        and (C,E) $V_M=0.0001$.
    }
    \label{fig:human-neand-h2}
\end{figure}

@@

This same result can be found by considering a haploid model
\citep{negm2024effect}. While the selection coefficient takes an equivalent
form to underdominance, we note that the mechanism differs: selection acts
against the minor allele, rather than for the heterozygous genotype.

@@

>>>>

    Many phenotypic traits are under stabilizing selection, which maintains a
    population's mean phenotypic value near some optimum. The dynamics of
    traits and trait architectures under stabilizing selection have bene
    extensively studied for single populations at steady state. However,
    natural populations are seldom at steady state and are often structured in
    some way. Admixture and introgression events may be common, including over
    human evolutionary history. Because stabilizing selection results in
    selection against the minor allele at a trait-affecting locus, alleles from
    the minor parental ancestry will be selected against after admixture. We
    show that the site-frequency spectrum can be used to model the genetic
    architecture of such traits, allowing for the study of trait architecture
    dynamics in complex multi-population settings. We use a simple
    deterministic two-locus model to predict the reduction of introgressed
    ancestry around trait-contributing loci. From this and individual-based
    simulations, we show that introgressed-ancestry deserts are enriched around
    such loci. When introgression between two diverged populations occurs in
    both directions, as has been inferred between humans and Neanderthals, the
    locations of introgressed-ancestry deserts will tend to be shared across
    populations. We argue that stabilizing selection for shared phenotypic
    optima may explain recent observations in which regions of depleted
    human-introgressed ancestry in the Neanderthal genome overlap with
    Neanderthal-ancestry deserts in humans.

@@

Once introduced through admixture, introgressed alleles may be selected for or
against. Some introgressed haplotypes were likely positively selected in modern
\emph{Homo sapiens} (here, ``humans'') \citep{huerta2014altitude,
racimo2017signatures, enard2018evidence, gower2021detecting}, possibly due to
locally adaptive variation that provided fitness advantages as humans
encountered novel environments. Despite some cases of adaptive introgression,
most introduced functional alleles likey were selected against in humans
\citep{harris2016genetic, juric2016strength, veller2023recombination}. Since
Neanderthal and Denisovan population sizes were relatively small for hundreds
of thousands of years, theory predicts they would have accumulated deleterious
variation at an increased rate. Introgressed haplotypes loaded with more
deleterious mutations would have been rapidly removed by selection after
admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
humans shows a reduction of Neanderthal-related ancestry in coding and
regulatory regions \citep{petr2019limits, telis2020selection,
yermakovich2023long}. These ``deserts'' of Neanderthal ancestry support the
hypothesis that introgressed functional alleles were selected against
\citep{sankararaman2014genomic, sankararaman2016combined}.

There is growing genetic evidence that Neanderthals reciprocally received
genetic material from early humans \citep{kuhlwilm2016ancient,
hubisz2020mapping, harris2023diverse, li2024recurrent}. This gene flow occurred
tens to hundreds of thousands of years prior to Neanderthal introgression in
humans during the global dispersal of modern humans around 60 ka. The earlier
admixture is supported by \emph{H. sapiens} outside of African around 200--100
ka \citep{schwarcz1988esr, grun2005u, harvati2019apidima, beyer2021climatic},
potentially overlapping with Neanderthals and providing opportunities for early
contacts. While estimates of the genomic contribution of early humans to
Neanderthals vary, around $6\%$ of later Neanderthal genomes may trace through
this admixture event \citep{harris2023diverse}. Under a ``load'' model, if
human-related haplotypes carried fewer deleterious alleles due to their larger
long-term effective population size, human-introgressed DNA would have been
favored in Neanderthal genomes. The replacement of Neanderthal mitochondrial
and Y chromosomes by early human haplotypes appears to support this model of
post-admixture positive selection in the Neanderthal lineage
\citep{posth2017deeply, petr2020evolutionary}.

Models for selection against introgressed alleles are often based on
deleterious load or hybrid incompatibilities \citep{muller1942isolating}. These
models, founded in population genetics theory, rarely take into account
selection operating on phenotypic traits or the relationship between genetic
and phenotypic variation.  Many phenotypic traits are thought to be under
stabilizing selection \citep{sanjak2018evidence, sella2019thinking}, including
gene regulation \citep{gilad2006natural, hodgins2015gene, price2022detecting}.
Because some of the strongest signals of selection against
Neanderthal-introduced alleles are in regulatory regions
\citep{sankararaman2014genomic}, stabilizing selection on quantitative traits
may be particularly relevant to the dynamics of functional genetic variation
after introgression among hominins.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/stab_sel.pdf}
    \caption{
        \textbf{Additive genetic variance under stabilizing selection.}
        (A) Stabilizing selection acts to maintain phenotypic values of
        individuals in the population near some optimum. Throughout, we assume
        a Gaussian fitness function.
        (B) With low mutational variance ($V_M$), the expected additive
        genetic variance ($V_G$) is proportional to the
        population-scaled mutation rate. When $V_M$ is large ($N_e V_M > 1$) so that
        mutational effects can be strong, $V_G$ is independent of $V_M$.
        The stochastic house-of-cards model (Eq.~\ref{eq:SHC})
        interpolates these regimes, assuming steady-state dynamics
        \citep{burger1989much}. Expected $V_G$ computed using an SFS approach
        \citep[developed here using \moments,][]{jouganous2017inferring}
        matches simulations assuming linkage equilibrium between loci affecting the
        trait, which align closely with the stochastic house-of-cards
        approximation.
        Here $V_S=1$, $N_e=10{,}000$, the mutation rate
        $\mu=0.01$ (per haploid), and mutation effects are drawn from a normal
        distribution with mean zero and variance $V_M$.
    }
    \label{fig:stab-sel}
\end{figure}

In diverged populations, the genetic variation contributing to a trait under
stabilizing selection has a higher rate of turnover compared to neutral
evolution \citep{yair2022population}. We therefore expect a rapid divergence of
trait architectures in the human and Neanderthal lineages at trait-contributing
loci, even when the mean phenotype in each population remains close to the same
trait optimum. When a derived allele at high frequency in one population is
introduced to another population in which it was previously absent, it will be
at low frequency (if the admixture proportion is low) and subsequently selected
against. Likewise, if the ancestral allele is reintroduced to a population
fixed for a derived allele, the ancestral allele will be at low frequency and
will be selected against. Selection acts against the introgressed allele in
either case, whether it is ancestral or derived and regardless of the
historical relative sizes of the populations involved. In concurrent work,
\citet{veller2024stabilizing} demonstrate this effect by deriving the expected
decay of minor parental ancestry under stabilizing selection after admixture.
This prediction contrasts with the population-genetics load model, in which
haplotypes with fewer deleterious variants (such as those from the population
with larger historical size) are favored after introgression in either
direction.

In this article, we show that admixture between diverged populations results in
selection against the minor parental ancestry at loci contributing to a trait
under stabilizing selection. We develop a numerical approach based on the
site-frequency spectrum to predict genetic variation under complex demographic
scenarios, which we use to partition predicted trait heritability by
introgressed and non-introgressed variation. Using simulations with linkage, we
demonstrate that deserts of introgressed ancestry form around
trait-contributing loci. When gene flow occurs bidirectionally, such deserts
will tend to overlap in location across populations. We argue that stabilizing
selection on shared trait optima may explain the overlap of
introgressed-ancestry deserts in human and Neanderthal genomes after reciprocal
introgression \citep{harris2023diverse}.

@@

We consider a polygenic trait for which an individual's additive genetic value
is the sum over all effects of alleles in their genome. For individual $i$,
\(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size of the derived
allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
locus (i.e., the number of derived alleles they carry). With linkage
equilibrium between trait-affecting loci, the expected additive genetic
variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
frequency at locus $l$. We ignore dominance and epistasis (often argued to be a
reasonable modeling choice, e.g., \citet{barton2002understanding,
zhang2002pleiotropic}), so that \(V_G = V_A\). We further ignore environmental
effects and noise \citep[under the assumption of additive environmental
effects, as in][]{lande1975maintenance, simons2018population}, so that only
genetic effects are considered (so that \(V_P=V_G\)).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$, typically set to zero, and we assume a Gaussian fitness function
(Figure~\ref{fig:stab-sel}A) so that relative fitness (relative to an
individual with optimal phenotype) is given by \(f(G_i | O,
V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $\frac{1}{V_S}$ is interpreted as the
strength of selection on the trait, so larger $V_S$ implies weaker selection.
For a population with mean phenotype at or very close to the optimum, the
mean fitness of the population [assuming a normal distribution of phenotypic
values in the population, \citep{turelli1994genetic, urban2013asymmetric}] is
\[\bar{w} \approx \int_{-\infty}^\infty
f(G | O, V_S) \mathcal{N}(O, V_G) dG =
\left(\frac{V_S}{V_S+V_G}\right)^{1/2},\]
so that as the genetic variance increases, mean fitness decreases.

@@

In our case, the selection coefficient $s$ depends on the strength of selection
on the trait $V_S$ and the effect size $a$, as well as the frequency of the
allele, so that
\begin{equation}\label{eq:selcoeff}
    s\approx -\frac{a^2(1-2p)}{2(V_S+V_G)} 
    = \frac{\left(p-\frac{1}{2}\right)a^2}{V_S + V_G}
    \approx \frac{\left(p-\frac{1}{2}\right)a^2}{V_S},
\end{equation}
when $V_G \ll V_S$. This model assumes linkage equilibrium between
trait-contributing loci. We note that LD is expected to be generated between
loci even in the absence of linkage \citep{bulmer1971effect}, and
\citet{negm2024effect} recently showed how to correct for its effect on the
selection dynamics of a focal allele.

@@

We extended the \moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017inferring} to include underdominance with given
selection coefficient $s$ (Appendix~\ref{sec:moments-underdominance}). The
contribution of alleles with effect size $a$ to the total genetic variance
$V_G$ is found by computing the expected pairwise diversity from the SFS (with
sample size $n$, denoted $\Phi_n$), as
\[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1} \frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a).\]
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
assuming a normal distribution of effect sizes for new mutations, the total
genetic variance is
\[V_G \approx \int_{-\infty}^\infty V_{G,a} dF(a),\]
where \(F(a)\) is the cumulative distribution function of \(\mathcal{N}(0,V_M)\).
$V_G$ can be computed periodically in this way to predict trajectories in
non-equilibrium settings (Figure~\ref{fig:one-pop}).

If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
$s=(p-1/2)a^2/V_S$ leads to deflated estimates of additive genetic variance
(Figure~\ref{fig:supp-one-pop}). When mutation rates are large so that $V_G$ is
not small compared to $V_S$, using $s=(p-1/2)a^2/(V_S+V_G)$ provides estimates
of $V_G$ that closely match simulations assuming linkage equilibrium between
loci (Figures~\ref{fig:toy-admixture} and
\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}). Because $V_G$ can vary over
time, this means that $s$ is no longer constant and can change due to factors
such as non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Population size changes and additive genetic variance.}
        We used \moments to track expected genetic variance of a trait
        under stabilizing selection as the population changes size. (A) The
        population goes through a 10-fold size reduction, followed by a recovery.
        (B) All mutations have effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
        probability of being trait increasing or decreasing. Depending on the
        selection coefficient (Equation~\ref{eq:selcoeff}) compared to $N_e$, $V_G$
        could increase or decrease after a sudden population size change.
        Predictions using \moments match simulations with linkage equilibrium
        between trait-affecting loci
        (Figures~\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}).
    }
    \label{fig:one-pop}
\end{figure}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model a population of size
$N_e=10{,}000$ splits into two, one remaining size $10{,}000$ and the other
shrinking to size $1{,}000$. They remain isolated for $2N_e$ generations (or
500 thousand years, assuming an average generation time of 25 years) and then
introgression occurs from one branch to the other, contributing $5\%$ ancestry
to the recipient population (Figure~\ref{fig:toy-admixture}A,B).

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/reciprocal_admixture.pdf}
    \caption{
        \textbf{Genetic variance of a trait under stabilizing selection with
        multi-population demography.}
        (A,B) Two simple demographic models, in which Deme 1 has size 10,000
        and Deme 2 has size 1,000. In each scenario, we allow $5\%$ admixture
        from one deme to the other after begin isolated for 20,000 ($2N_e$)
        generations.
        (C,D) We compared predicted (additive) $V_G$ from the site-frequency
        spectra (using \moments) to simulations
        without linkage between trait-affecting loci. In Deme 2, $V_G$ decreases
        due to their population size reduction. After admixture in both
        directions, $V_G$ increases in the recipient deme
        and then decays to steady-state levels. Predictions from \moments closely
        match observed $V_G$ in simulations.
        Here, mutations were drawn from a normal distribution with $V_M=0.0025$.
        Other parameters: $\mu=0.025$, $V_S=1$, and the optimal phenotype remained
        the same in each population.
        (E,F) Using \moments, we partitioned additive contributions to $V_G$ by
        mutations that were previously segregating in the focal population,
        introgressed variants (either derived or reintroduced ancestral alleles),
        and by new mutations since the time of admixture. In both demographic
        scenarios, introgressed variants contribute a substantial proportion to
        $V_G$, though it is primarily composed of introduced derived alleles when
        admixture is from the small to the large population, while primarily
        reintroduced ancestral alleles under the reverse direction of gene flow.
        In both cases, $V_G$ is increasingly due to new mutations as the genetic
        architecture of the trait turns over with time since admixture.
    }
    \label{fig:toy-admixture}
\end{figure}

@@

We compared our \moments-based predictions for $V_G$ in non-equilibrium
settings to simulations assuming linkage equilibrium as well as
individual-based simulations with recombination (next section). For both,
mutations occur at rate $\mu$ per haploid genome copy per generation, with
effects drawn from a normal distribution with mean 0 and variance $V_M$.

@@

We typically expect genetic variance to increase after introgression. The
amount that genetic variance increases depends on the allelic differences
accumulated between populations and the effects of those alleles. Assuming
linkage equilibrium, and ignoring dominance and epistasis, \(V_G=\sum_l
2p_l(1-p_l)a_l^2\). After admixture, with proportion $f$ contributed by the
population labeled 0 and $1-f$ by population 1, \(p_l = fp_{l,0} +
(1-f)p_{l,1}\). Plugging into the expression for $V_G$ and after a bit of
algebra (Appendix~\ref{sec:VG-admixture}), we can write the expected genetic
variance directly after admixture as
\begin{align}
V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,
\end{align}
where $F_2 = (p_0 - p_1)^2$ is the squared difference in allele frequencies at
a locus \citep{peter2016admixture}. This result is known
\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
to that in the source populations weighted by their contributions, plus a term
that depends on the divergence at trait-affecting loci between the populations
weighted by the quadratic factor $2f(1-f)$. We note that $V_G$ is expected to
increase only in the second generation after admixture, so that selection
against introgressed alleles is not immediate \citep{veller2024stabilizing}.
This effect is not captured by this expression.

$F_2$ at a given locus depends on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on demographic history. In this case,
\begin{align} \label{eq:VG-admixture}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    \nonumber
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
\end{align}
where $L$ is the number of trait affecting loci.

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/h2-per-SNP.pdf}
    \caption{
        \textbf{Allele contributions to heritability under human-Neanderthal
        reciprocal introgression.}
        (A) A model of divergence and admixture between humans and
        Neanderthals. Using \moments, we computed predicted $V_G$ over time,
        partitioned by variation that was introgressed vs. non-introgressed
        (Figure~\ref{fig:human-neand-VG}).
        (B-E) Predicted per-SNP contributions to genetic variance ($h^2$ per
        SNP) is plotted over the 50 thousand years following introgression.
        For non-introgressed variants, we also plot $h^2$-per-SNP weighted by
        allele frequencies matching those of introgressed variants. These
        are shown for (B,C) Neanderthal-to-human introgression 50 kya,
        (D,E) human-to-Neanderthal introgression 250 kya, (B,D) $V_M=0.0025$,
        and (C,E) $V_M=0.0001$.
    }
    \label{fig:human-neand-h2}
\end{figure}

@@

This same result can be found by directly considering the marginal fitness of a
focal allele against the expected haploid background, as pointed out by
\citet{negm2024effect}. Thus, while the selection coefficient takes an
equivalent form to underdominance, we note that the mechanism differs:
selection acts against the minor allele, rather than against the heterozygous
genotype.

@@

\hfill$\square$
